---
title: "Impact of Covid-19 on Vessel densities in the Belgian Part of the North Sea"
author: "Lennert Schepers, Flanders Marine Data Centre (VMDC), VLIZ"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this exercise we want to analyse the impact of COVID-19 on vessel densities in the Belgian Part of the North Sea and the Western Scheldt.
There are different ways to calculate vessel densities, therefore we first start with some background information on two different types of vessel densities and where this data can be found.
The data and code of this analysis if freely available, see the 'Data availability' and 'Code availability' sections at the end of this document.

## Vessel densities

In this small exercise we want to analyse vessel route densities in the Belgian part of the North Sea.
The data for this exercise comes from [EMODnet Human activities](https://www.emodnet-humanactivities.eu/).
EMODnet Human Activities has two types of vessel density data, one created by the Human Activities portal themselves, giving the vessel hours per square km per month by ship type. See [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Vessel+Density+).

And one created by the European Maritime Safety Agency (EMSA), giving the number of routes per square km per month) by ship type. The advantage of the latter is that this provides recent information. For example, writing now 28th of April, the monthly aggregated data is already available for March 2020. For details, see [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Route+density+%28source%3A+EMSA%29) 


```{r libs, message = FALSE, warning = FALSE, echo = FALSE}
# First, we load the R libraries that we will use
library(raster)  # for raster data
library(sf)      # for spatial data
library(mapview) # for interactive maps
library(ggplot2) # for plots
library(data.table) # for manipulation dataframes/tables
library(DT)
library(mregions)  # for standardize marine regions from http://www.marineregions.org
```

The focus of this exercise is the Belgian waters (the Belgian Exclusive Economic Zone) and the Western Scheldt
```{r message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
# BCP <- mr_shp(key = "MarineRegions:eez",
#                   filter = "Belgian Exclusive Economic Zone",
#                   maxFeatures = 500) # needed for CRAN version, see https://github.com/ropensci/mregions/issues/48
BCP <- st_read("http://geo.vliz.be/geoserver/MarineRegions/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=eez&cql_filter=mrgid=%273293%27&outputFormat=application/json")
# BCP <- st_cast(BCP, "POLYGON")
BCP_bbox <- st_bbox(BCP$geometry)


B3NM <- st_read("http://geo.vliz.be/geoserver/Belgium/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=be_3nm_2015&outputFormat=application/json")

B12NM <- st_read("http://geo.vliz.be/geoserver/Belgium/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=Belgium:be_12nm_2015&outputFormat=application/json")


mapview(BCP, color = 'blue', alpha.regions = 0) +
  mapview(B3NM) +
  mapview(B12NM)

st_write(B3NM, file.path('data', 'B3NM.gpkg'), delete_layer = TRUE)
st_write(B12NM, file.path('data', 'B12NM.gpkg'), delete_layer = TRUE)
st_write(BCP, file.path('data', 'BCP.gpkg'), delete_layer = TRUE)
```

```{r message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
WScheldt <- st_read("http://geo.vliz.be/geoserver/wfs?request=getfeature&service=wfs&version=1.1.0&typename=MarineRegions:seavox_v16&outputformat=json&filter=%3CPropertyIsEqualTo%3E%3CPropertyName%3Esub_region%3C%2FPropertyName%3E%3CLiteral%3EWESTERN+SCHELDT%3C%2FLiteral%3E%3C%2FPropertyIsEqualTo%3E")
WScheldt_bbox <- st_bbox(WScheldt$geometry)
#mapview(WScheldt)

bbox_BCP_Sch <- st_bbox(st_union(BCP,WScheldt))

st_write(WScheldt, file.path('data', 'WScheldt.gpkg'), delete_layer = TRUE)

```




```{r WCS function, message = FALSE, warning = FALSE, echo = FALSE}
# Now, we create a function to retrieve the route density for a particular fishing type, month and year *in Belgian waters*

base_url <- 'https://ows.emodnet-humanactivities.eu/wcs?SERVICE=WCS&VERSION=1.0.0&request=GetCapabilities'

# BACKGROUND INFO ON EMODNet Human Activities' WCS Service, and the 'route density maps'
    # The Open Geospatial Consortium (OGC) Web Coverage Service Interface Standard (WCS) defines Web-based retrieval of coverages – that is, digital geospatial information representing space/time-varying phenomena.
    #
    # Get Capabilities:
    #   https://ows.emodnet-humanactivities.eu/wcs?SERVICE=WCS&VERSION=1.0.0&request=GetCapabilities
    #
    # Get Coverage Example:
    #   https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=getcoverage&coverage=emodnet:2019_01_rd_All&crs=EPSG:4326&BBOX=15,20.5,30,32.5&format=image/tiff&interpolation=nearest&resx=0.00833333&resy=0.00833333
    #
    # The names of the web service layers are -
    #   '<year>_<month>_rd_<vessel code>' Eg. '2019_01_rd_All' for all vessels during January 2019
    # '<year>_<season>_rd_<vessel code>' Eg. '2019_spring_rd_All' for all vessels during Spring 2019
    # '<year>_yearly_rd_<vessel code>' Eg. '2019_yearly_rd_All' for the total of all vessels during 2019.
    # The vessel codes are as follows:
    #
    # All - All types
    # 01 - Cargo
    # 02 - Fishing
    # 03 - Passenger
    # 04 - Tanker
    # 05 - Other
    #
    #Available data: January 2019 - March 2020

# function to get belgian raster data, based on year, month, vessel_code
getBelgianCSW <- function(year, month, vessel_code){
  url <- paste0("https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=getcoverage&coverage=emodnet:",
                year, "_",
                formatC(month, width = 2, flag = 0), "_rd_",
                vessel_code, "&",
                "crs=EPSG:4326&BBOX=", paste(bbox_BCP_Sch, collapse = ","), "&",
                "format=image/tiff&",
                "interpolation=nearest&resx=0.00833333&resy=0.00833333"
  )
  x <- raster(url)
}

```

We extract the Vessel density data, for example here for the Fishing map of January 2019:

```{r get_WCS_data, message = FALSE, warning = FALSE, echo = FALSE}
# extract vessel density data and store it in a rasterbrick:

Cargo <- brick(x = c('Cargo-2019-01' = getBelgianCSW(2019, 1, "01"),
                     'Cargo-2019-02' = getBelgianCSW(2019, 2, "01"),
                     'Cargo-2019-03' = getBelgianCSW(2019, 3, "01"),
                     'Cargo-2019-04' = getBelgianCSW(2019, 4, "01"),
                     'Cargo-2019-05' = getBelgianCSW(2019, 5, "01"),
                     'Cargo-2019-06' = getBelgianCSW(2019, 6, "01"),
                     'Cargo-2019-07' = getBelgianCSW(2019, 7, "01"),
                     'Cargo-2019-08' = getBelgianCSW(2019, 8, "01"),
                     'Cargo-2019-09' = getBelgianCSW(2019, 9, "01"),
                     'Cargo-2019-10' = getBelgianCSW(2019, 10, "01"),
                     'Cargo-2019-11' = getBelgianCSW(2019, 11, "01"),
                     'Cargo-2019-12' = getBelgianCSW(2019, 12, "01"),
                     'Cargo-2020-01' = getBelgianCSW(2020, 1, "01"),
                     'Cargo-2020-02' = getBelgianCSW(2020, 2, "01"),
                     'Cargo-2020-03' = getBelgianCSW(2020, 3, "01"),
                     'Cargo-2020-04' = getBelgianCSW(2020, 4, "01")
                                   )
)


Fishing <- brick(x = c('Fishing-2019-01' = getBelgianCSW(2019, 1, "02"),
                     'Fishing-2019-02' = getBelgianCSW(2019, 2, "02"),
                     'Fishing-2019-03' = getBelgianCSW(2019, 3, "02"),
                     'Fishing-2019-04' = getBelgianCSW(2019, 4, "02"),
                     'Fishing-2019-05' = getBelgianCSW(2019, 5, "02"),
                     'Fishing-2019-06' = getBelgianCSW(2019, 6, "02"),
                     'Fishing-2019-07' = getBelgianCSW(2019, 7, "02"),
                     'Fishing-2019-08' = getBelgianCSW(2019, 8, "02"),
                     'Fishing-2019-09' = getBelgianCSW(2019, 9, "02"),
                     'Fishing-2019-10' = getBelgianCSW(2019, 10, "02"),
                     'Fishing-2019-11' = getBelgianCSW(2019, 11, "02"),
                     'Fishing-2019-12' = getBelgianCSW(2019, 12, "02"),
                     'Fishing-2020-01' = getBelgianCSW(2020, 1, "02"),
                     'Fishing-2020-02' = getBelgianCSW(2020, 2, "02"),
                     'Fishing-2020-03' = getBelgianCSW(2020, 3, "02"),
                     'Fishing-2020-04' = getBelgianCSW(2020, 4, "02")
                     )
)

mapview(Fishing[[1]])

Passenger <- brick(x = c('Passenger-2019-01' = getBelgianCSW(2019, 1, "03"),
                        'Passenger-2019-02' = getBelgianCSW(2019, 2, "03"),
                        'Passenger-2019-03' = getBelgianCSW(2019, 3, "03"),
                        'Passenger-2019-04' = getBelgianCSW(2019, 4, "03"),
                        'Passenger-2019-05' = getBelgianCSW(2019, 5, "03"),
                        'Passenger-2019-06' = getBelgianCSW(2019, 6, "03"),
                        'Passenger-2019-07' = getBelgianCSW(2019, 7, "03"),
                        'Passenger-2019-08' = getBelgianCSW(2019, 8, "03"),
                        'Passenger-2019-09' = getBelgianCSW(2019, 9, "03"),
                        'Passenger-2019-10' = getBelgianCSW(2019, 10, "03"),
                        'Passenger-2019-11' = getBelgianCSW(2019, 11, "03"),
                        'Passenger-2019-12' = getBelgianCSW(2019, 12, "03"),
                        'Passenger-2020-01' = getBelgianCSW(2020, 1, "03"),
                        'Passenger-2020-02' = getBelgianCSW(2020, 2, "03"),
                        'Passenger-2020-03' = getBelgianCSW(2020, 3, "03"),
                        'Passenger-2020-04' = getBelgianCSW(2020, 4, "03")
)
)

Tanker <- brick(x = c('Tanker-2019-01' = getBelgianCSW(2019, 1, "04"),
                      'Tanker-2019-02' = getBelgianCSW(2019, 2, "04"),
                      'Tanker-2019-03' = getBelgianCSW(2019, 3, "04"),
                      'Tanker-2019-04' = getBelgianCSW(2019, 4, "04"),
                      'Tanker-2019-05' = getBelgianCSW(2019, 5, "04"),
                      'Tanker-2019-06' = getBelgianCSW(2019, 6, "04"),
                      'Tanker-2019-07' = getBelgianCSW(2019, 7, "04"),
                      'Tanker-2019-08' = getBelgianCSW(2019, 8, "04"),
                      'Tanker-2019-09' = getBelgianCSW(2019, 9, "04"),
                      'Tanker-2019-10' = getBelgianCSW(2019, 10, "04"),
                      'Tanker-2019-11' = getBelgianCSW(2019, 11, "04"),
                      'Tanker-2019-12' = getBelgianCSW(2019, 12, "04"),
                      'Tanker-2020-01' = getBelgianCSW(2020, 1, "04"),
                      'Tanker-2020-02' = getBelgianCSW(2020, 2, "04"),
                      'Tanker-2020-03' = getBelgianCSW(2020, 3, "04"),
                      'Tanker-2020-04' = getBelgianCSW(2020, 4, "04")
)
)

Other <- brick(x = c('Other-2019-01' = getBelgianCSW(2019, 1, "05"),
                      'Other-2019-02' = getBelgianCSW(2019, 2, "05"),
                      'Other-2019-03' = getBelgianCSW(2019, 3, "05"),
                      'Other-2019-04' = getBelgianCSW(2019, 4, "05"),
                      'Other-2019-05' = getBelgianCSW(2019, 5, "05"),
                      'Other-2019-06' = getBelgianCSW(2019, 6, "05"),
                      'Other-2019-07' = getBelgianCSW(2019, 7, "05"),
                      'Other-2019-08' = getBelgianCSW(2019, 8, "05"),
                      'Other-2019-09' = getBelgianCSW(2019, 9, "05"),
                      'Other-2019-10' = getBelgianCSW(2019, 10, "05"),
                      'Other-2019-11' = getBelgianCSW(2019, 11, "05"),
                      'Other-2019-12' = getBelgianCSW(2019, 12, "05"),
                      'Other-2020-01' = getBelgianCSW(2020, 1, "05"),
                      'Other-2020-02' = getBelgianCSW(2020, 2, "05"),
                      'Other-2020-03' = getBelgianCSW(2020, 3, "05"),
                      'Other-2020-04' = getBelgianCSW(2020, 4, "05")
)
)

All <- brick(x = c('All-2019-01' = getBelgianCSW(2019, 1, "All"),
                   'All-2019-02' = getBelgianCSW(2019, 2, "All"),
                   'All-2019-03' = getBelgianCSW(2019, 3, "All"),
                   'All-2019-04' = getBelgianCSW(2019, 4, "All"),
                   'All-2019-05' = getBelgianCSW(2019, 5, "All"),
                   'All-2019-06' = getBelgianCSW(2019, 6, "All"),
                   'All-2019-07' = getBelgianCSW(2019, 7, "All"),
                   'All-2019-08' = getBelgianCSW(2019, 8, "All"),
                   'All-2019-09' = getBelgianCSW(2019, 9, "All"),
                   'All-2019-10' = getBelgianCSW(2019, 10, "All"),
                   'All-2019-11' = getBelgianCSW(2019, 11, "All"),
                   'All-2019-12' = getBelgianCSW(2019, 12, "All"),
                   'All-2020-01' = getBelgianCSW(2020, 1, "All"),
                   'All-2020-02' = getBelgianCSW(2020, 2, "All"),
                   'All-2020-03' = getBelgianCSW(2020, 3, "All"),
                   'All-2020-04' = getBelgianCSW(2020, 4, "All")
)
)

```

```{r writeraster, message = FALSE, warning = FALSE, echo = FALSE}
if (!dir.exists("output")) dir.create("output")

names <- c('Cargo', 'Fishing', 'Passenger', 'Tanker', 'Other', 'All')
bricks <- c(Cargo, Fishing, Passenger, Tanker, Other, All)

l <- lapply(bricks,
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)

```



### Belgian EEZ:

We extract all data from all the cells, and calculate the average for the whole BCP.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# No we create a data frame to store the statistics that we will extract from the rasters.
# We extract for all values of the Belgian EEZ, the average value of all raster cells.

# create data frame, with 1st column 'time'
  df.sr <- data.frame(Date = seq(as.Date("2019/1/1"), by = "month", length.out = 16))

  # extract all values from the rasters, by the BCP polygon, and average the values.
  df.sr$Cargo     <- as.vector(extract(Cargo, BCP, fun = mean))
  df.sr$Fishing   <- as.vector(extract(Fishing, BCP, fun = mean))
  df.sr$Passenger <- as.vector(extract(Passenger, BCP, fun = mean))
  df.sr$Tanker    <- as.vector(extract(Tanker, BCP, fun = mean))
  df.sr$Other     <- as.vector(extract(Other, BCP, fun = mean))
  df.sr$All       <- as.vector(extract(All, BCP, fun = mean))

# transform to long format:
  df.sr.long <- melt.data.table(as.data.table(df.sr),
                                id.vars = "Date")
```
This is an example of the data in 'wide' format

```{r BCP_table_wide, echo = FALSE}
datatable(df.sr,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)
```

```{r BCP_table_long, echo = FALSE, eval = FALSE}
datatable(df.sr.long,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(3, 2)
```

A plot of the result:

```{r plot, message = FALSE, warning = FALSE, echo = FALSE}

ggplot(df.sr.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Vessel route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()
```


### Scheldt estuary:

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# create data frame, with 1st column 'time'
  df.Scheldt <- data.frame(Date = seq(as.Date("2019/1/1"), by = "month", length.out = 16))

  # extract all values from the rasters, by the WScheldt polygon, and average the values.
  df.Scheldt$Cargo     <- as.vector(extract(Cargo, WScheldt, fun = mean))
  df.Scheldt$Fishing   <- as.vector(extract(Fishing, WScheldt, fun = mean))
  df.Scheldt$Passenger <- as.vector(extract(Passenger, WScheldt, fun = mean))
  df.Scheldt$Tanker    <- as.vector(extract(Tanker, WScheldt, fun = mean))
  df.Scheldt$Other     <- as.vector(extract(Other, WScheldt, fun = mean))
  df.Scheldt$All       <- as.vector(extract(All, WScheldt, fun = mean))

# transform to long format:
  df.Scheldt.long <- melt.data.table(as.data.table(df.Scheldt),
                                id.vars = "Date")

```

```{r Scheldt table, echo = FALSE}
datatable(df.Scheldt,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)
```


A plot of the result:

```{r Schelde_plot, message = FALSE, warning = FALSE, echo = FALSE}

ggplot(df.Scheldt.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Vessel route densities in the Western Scheldt",
       subtitle = "Monthly average route density (routes/km²/ship type) in the Western Scheldt",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()
```


## 2020 vs 2019

We create a mean value for

 - February - April 2019
 - February - April 2020
 
And plot both periods rasters
(the map is the Fishing type example)

```{r message = FALSE, warning = FALSE, echo = FALSE}
# Average map of February-April 2019
r2019_0204 <- brick(c(
  Cargo_2019_0204 = mean(Cargo[[2:4]]),
  Fishing_2019_0204 = mean(Fishing[[2:4]]),
  Passenger_2019_0204 = mean(Passenger[[2:4]]),
  Tanker_2019_0204 = mean(Tanker[[2:4]]),
  Other_2019_0204 = mean(Other[[2:4]]),
  All_2019_0204 = mean(All[[2:4]])
)
)
  
# Average map of February-April 2020
r2020_0204 <- brick(c(
  Cargo_2020_0204 = mean(Cargo[[14:16]]),
  Fishing_2020_0204 = mean(Fishing[[14:16]]),
  Passenger_2020_0204 = mean(Passenger[[14:16]]),
  Tanker_2020_0204 = mean(Tanker[[14:16]]),
  Other_2020_0204 = mean(Other[[14:16]]),
  All_2020_0204 = mean(All[[14:16]])
)
)


l <- lapply(list(r2019_0204,r2020_0204),
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)

```

```{r mapviewpolygons, message = FALSE, warning = FALSE, echo = FALSE}
mapviewpolygons <- 
  mapview(BCP, color = 'blue',
        alpha.regions = 0,
        legend = FALSE,
        label = NULL,
        popup = NULL,
        homebutton = FALSE) +
  mapview(WScheldt,
          color = 'blue',
          alpha.regions = 0,
          legend = FALSE,
          label = NULL,
          popup = NULL,
        homebutton = FALSE) +
  mapview(B3NM,
          legend = FALSE, 
          label = NULL,
          popup = NULL,
        homebutton = FALSE) +
  mapview(B12NM,
          legend = FALSE, 
          label = NULL,
          popup = NULL,
        homebutton = FALSE)

```


### Cargo 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Cargo_2019_0204) + 
  mapview(r2020_0204$Cargo_2020_0204) 

```

### Fishing 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Fishing_2019_0204) + 
  mapview(r2020_0204$Fishing_2020_0204)
```

### Passenger 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020
mapviewpolygons +
  mapview(r2019_0204$Passenger_2019_0204) + 
  mapview(r2020_0204$Passenger_2020_0204)
```

### Tanker 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020
mapviewpolygons +
  mapview(r2019_0204$Tanker_2019_0204) + 
  mapview(r2020_0204$Tanker_2020_0204)
```

### Other 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Other_2019_0204) + 
  mapview(r2020_0204$Other_2020_0204)
```

### All 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$All_2019_0204) + 
  mapview(r2020_0204$All_2020_0204)
```


## Anomaly maps

### February - April 2019 compared to February - April 2020
Comparing the period February-April 2020 vs 2019:
(the different layers are the different boat types)


```{r echo = FALSE}
anomaly_February_April <- r2020_0204 - r2019_0204
names(anomaly_February_April) <- c("Anomaly_Cargo_Feb_Apr_2019_2020",
                                   "Anomaly_Fishing_Feb_Apr_2019_2020",
                                   "Anomaly_Passenger_Feb_Apr_2019_2020",
                                   "Anomaly_Tanker_Feb_Apr_2019_2020",
                                   "Anomaly_Other_Feb_Apr_2019_2020",
                                   "Anomaly_All_Feb_Apr_2019_2020") 
writeRaster(anomaly_February_April,
            gsub('\\.','_',file.path('output', names(anomaly_February_April))),
            format = "GTiff",
            bylayer = TRUE,
            overwrite = TRUE)


pal <- grDevices::colorRampPalette(grDevices::hcl.colors(255, palette = "BrBG"))

mapviewpolygons +
  mapView(anomaly_February_April,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )

```


### Month-by-month comparisons 2019 vs 2020
Comparing the month-by-month differences between 2020 vs 2019:
(the map is the Fishing type example)

```{r message = FALSE, warning = FALSE, echo = FALSE}
# Anomaly map January-April 2020 - January-April 2019
Anomaly_Cargo <- brick(c(
  Anomaly_Cargo_January = mean(Cargo[[13]] - Cargo[[1]]),
  Anomaly_Cargo_February = mean(Cargo[[14]] - Cargo[[2]]),
  Anomaly_Cargo_March = mean(Cargo[[15]] - Cargo[[3]]),
  Anomaly_Cargo_April = mean(Cargo[[16]] - Cargo[[4]])
)
)

Anomaly_Fishing <- brick(c(
  Anomaly_Fishing_January = mean(Fishing[[13]] - Fishing[[1]]),
  Anomaly_Fishing_February = mean(Fishing[[14]] - Fishing[[2]]),
  Anomaly_Fishing_March = mean(Fishing[[15]] - Fishing[[3]]),
  Anomaly_Fishing_April = mean(Fishing[[16]] - Fishing[[4]])
)
)

Anomaly_Passenger <- brick(c(
  Anomaly_Passenger_January = mean(Passenger[[13]] - Passenger[[1]]),
  Anomaly_Passenger_February = mean(Passenger[[14]] - Passenger[[2]]),
  Anomaly_Passenger_March = mean(Passenger[[15]] - Passenger[[3]]),
  Anomaly_Passenger_April = mean(Passenger[[16]] - Passenger[[4]])
)
)

Anomaly_Tanker <- brick(c(
  Anomaly_Tanker_January = mean(Tanker[[13]] - Tanker[[1]]),
  Anomaly_Tanker_February = mean(Tanker[[14]] - Tanker[[2]]),
  Anomaly_Tanker_March = mean(Tanker[[15]] - Tanker[[3]]),
  Anomaly_Tanker_April = mean(Tanker[[16]] - Tanker[[4]])
)
)

Anomaly_Other <- brick(c(
  Anomaly_Other_January = mean(Other[[13]] - Other[[1]]),
  Anomaly_Other_February = mean(Other[[14]] - Other[[2]]),
  Anomaly_Other_March = mean(Other[[15]] - Other[[3]]),
  Anomaly_Other_April = mean(Other[[16]] - Other[[4]])
)
)

Anomaly_All <- brick(c(
  Anomaly_All_January = mean(All[[13]] - All[[1]]),
  Anomaly_All_February = mean(All[[14]] - All[[2]]),
  Anomaly_All_March = mean(All[[15]] - All[[3]]),
  Anomaly_All_April = mean(All[[16]] - All[[4]])
)
)


l <- lapply(list(Anomaly_Cargo, Anomaly_Fishing, Anomaly_Passenger,
            Anomaly_Tanker, Anomaly_Other, Anomaly_All
            ),
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)


```

### Cargo 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Cargo,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Fishing 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Fishing,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Passenger 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Passenger,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Tanker 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Tanker,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Other 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Other,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### All 
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_All,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

## Additional analyses

### Passengers excluding wind farm

There is quite some traffic to the windfarms that are classified as 'Passenger' traffic.
As this might be confusing, we will exclude this from the analysis.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
passenger_windfarm <- st_read('data/passenger_windfarm.gpkg')

mapviewpolygons +
  mapview(passenger_windfarm) +
  mapview(Anomaly_Passenger,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )

```

```{r, message = FALSE, warning = FALSE, echo = FALSE}

BCP_no_windfarm <- st_difference(BCP, passenger_windfarm)

df.sr$Pass_no_windfarm <- as.vector(extract(Passenger, BCP_no_windfarm, fun = mean))

df.sr.pass <- df.sr[c('Date','Passenger','Pass_no_windfarm')]

# transform to long format:
  df.sr.pass.long <- melt.data.table(as.data.table(df.sr.pass),
                                id.vars = "Date")


```

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
datatable(df.sr.pass,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)

ggplot(df.sr.pass.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Passenger route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()

```

### Fishing zones.

In this part we'll look how much is being fished in the different zones:

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
# extended the lines of 3 and 12 NM a bit
BCP_zones <- st_read('data/eez_be_poly_split.gpkg', quiet = TRUE)
BCP_zones$name <- c(">12 NM", "0-3 NM", "3-12 NM")
mapview(BCP_zones)

#feb-apr20 vs feb-apr19 Fishing
numberofcells <- extract(Fishing, BCP_zones,
                         fun = function(x, ...) length(x),
                         df = TRUE)

df.Fishing.zones <- extract(Fishing, BCP_zones, fun = mean, df = TRUE)
df.Fishing.zones <- transpose(df.Fishing.zones[-1])
colnames(df.Fishing.zones) <- BCP_zones$name
df.Fishing.zones$Date <- seq(as.Date("2019/1/1"), by = "month", length.out = 16)

df.Fishing.zones <- df.Fishing.zones[c("Date",  "0-3 NM", "3-12 NM", ">12 NM")]

# transform to long format:
  df.Fishing.zones.long <- melt.data.table(as.data.table(df.Fishing.zones),
                                id.vars = "Date")


```

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
datatable(df.Fishing.zones,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)

ggplot(df.Fishing.zones.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Fishing route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "zone",
       shape = "zone") +
  theme_minimal()


```


## Data availability

- *Vessel densities*
The data from this exercise is freely available at the [EMODnet Human activities](https://www.emodnet-humanactivities.eu/) portal. [EMODnet](https://www.emodnet.eu/) stands for the European Marine Observation and Data Network, EMODnet is a network of organisations that are collecting, standardizing and making European marine data and data products freely available, supported by EU's integrated marine policy.
The vessel densities used in this exercise is provided by the European Maritime Safety Agency (EMSA) to EMODnet human activities and available [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Vessel+Density+).

- *Maritime boundaries*
The maritime boundaries used in this exercise are from [MarineRegions.org](http://www.marineRegions.org). MarineRegions.org maintains a standard, relational list of geographic names coupled with information and maps of the geographic location of these features. This improves access and clarity of the different geographic, marine names and allows an improved linking of these locations to databases. Marine Regions is devloped by Flanders Marine Institute (VLIZ) as part of the Flemish contribution to [LifeWatch](http://www.lifewatch.be/), funded by [Research Foundation - Flanders](https://www.fwo.be/en/).
The more information about the polygons used in this exercise:
  - the Western Scheldt: http://marineregions.org/gazetteer.php?p=details&id=4752
  - the Belgian Exclusive Economic Zone: http://marineregions.org/gazetteer.php?p=details&id=3293

## Code availability

All the code needed to run this analysis is available [here](index.Rmd).

This code makes use of following R packages:

- raster: for raster data
- sf: for spatial data
- mapview: for interactive maps
- ggplot2: for plots
- data.table: for manipulation dataframes/tables
- DT: for visualisation of the data tables
- mregions: for standardize marine regions from http://www.marineregions.org

