---
title: "Impact of Covid-19 on Vessel densities in the Belgian Part of the North Sea"
author: "Analysis by Flanders Marine Data Centre (VMDC), VLIZ"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this exercise we want to analyse the impact of COVID-19 on vessel densities in the Belgian Part of the North Sea and the Western Scheldt.
There are different ways to calculate vessel densities, therefore we first start with some background information on two different types of vessel densities and where this data can be found.
The data and code of this analysis are freely available, see the 'Data availability' and 'Code availability' sections at the end of this document. You can even reproduce the full analysis or modify parts of it, see the 'reproducibility' section.

This analysis is used in following policy information note (BIN) from Flanders Marine Institute:
http://www.vliz.be/nl/catalogus?module=ref&refid=324473


## Vessel densities

In this exercise we want to analyse vessel route densities in the Belgian part of the North Sea.
The data for this exercise comes from [EMODnet Human activities](https://www.emodnet-humanactivities.eu/).
EMODnet Human Activities has two types of vessel density data, one created by the Human Activities portal themselves, giving the vessel hours per square km per month by ship type. See [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Vessel+Density+).

And one created by the European Maritime Safety Agency (EMSA), giving the number of routes per square km per month) by ship type. The advantage of the latter is that this provides recent information, montly data is available with only 2-3 weeks delay.For details, see [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Route+density+%28source%3A+EMSA%29). We will use the EMSA dataset in the next sections.

All units are **number of routes / km² / ship type**


```{r libs, message = FALSE, warning = FALSE, echo = FALSE}
# First, we load the R libraries that we will use
library(raster)  # for raster data
library(sf)      # for spatial data
library(mapview) # for interactive maps
library(ggplot2) # for plots
library(data.table) # for manipulation dataframes/tables
library(DT)
library(mregions)  # for standardize marine regions from http://www.marineregions.org
```

The focus of this exercise is the Belgian waters (the Belgian Exclusive Economic Zone, EEZ) and the Western Scheldt
```{r message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
# BCP <- mr_shp(key = "MarineRegions:eez",
#                   filter = "Belgian Exclusive Economic Zone",
#                   maxFeatures = 500) # needed for CRAN version, see https://github.com/ropensci/mregions/issues/48
BCP <- st_read("http://geo.vliz.be/geoserver/MarineRegions/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=eez&cql_filter=mrgid=%273293%27&outputFormat=application/json")
# BCP <- st_cast(BCP, "POLYGON")
BCP_bbox <- st_bbox(BCP$geometry)


B3NM <- st_read("http://geo.vliz.be/geoserver/Belgium/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=be_3nm_2015&outputFormat=application/json")

B12NM <- st_read("http://geo.vliz.be/geoserver/Belgium/wfs?service=WFS&version=2.0.0&request=GetFeature&typeNames=Belgium:be_12nm_2015&outputFormat=application/json")


mapview(BCP, color = 'blue', alpha.regions = 0) +
  mapview(B3NM) +
  mapview(B12NM)

if (!dir.exists("data")) dir.create("data")
st_write(B3NM, file.path('data', 'B3NM.gpkg'), delete_layer = TRUE)
st_write(B12NM, file.path('data', 'B12NM.gpkg'), delete_layer = TRUE)
st_write(BCP, file.path('data', 'BCP.gpkg'), delete_layer = TRUE)
```

```{r message = FALSE, warning = FALSE, echo = FALSE, results = 'hide'}
WScheldt <- st_read("http://geo.vliz.be/geoserver/wfs?request=getfeature&service=wfs&version=1.1.0&typename=MarineRegions:seavox_v16&outputformat=json&filter=%3CPropertyIsEqualTo%3E%3CPropertyName%3Esub_region%3C%2FPropertyName%3E%3CLiteral%3EWESTERN+SCHELDT%3C%2FLiteral%3E%3C%2FPropertyIsEqualTo%3E")
WScheldt_bbox <- st_bbox(WScheldt$geometry)
#mapview(WScheldt)

bbox_BCP_Sch <- st_bbox(st_union(BCP,WScheldt))

st_write(WScheldt, file.path('data', 'WScheldt.gpkg'), delete_layer = TRUE)

```




```{r WCS function, message = FALSE, warning = FALSE, echo = FALSE}
# Now, we create a function to retrieve the route density for a particular fishing type, month and year *in Belgian waters*

# BACKGROUND INFO ON EMODNet Human Activities' WCS Service, and the 'route density maps'
# GO to https://www.emodnet-humanactivities.eu/search-results.php?dataname=Vessel+Density+

# create data/raster dir
if(!dir.exists(file.path('data','raster'))) dir.create(file.path('data','raster'))

# function to get belgian raster data, based on year, month, vessel_code
getBelgianCSW <- function(year, month, vessel_code){
  url <- paste0("https://ows.emodnet-humanactivities.eu/wcs?service=wcs&version=1.0.0&request=getcoverage&coverage=emodnet:",
                year, "_",
                formatC(month, width = 2, flag = 0), "_rd_",
                vessel_code, "&",
                "crs=EPSG:4326&BBOX=", paste(bbox_BCP_Sch, collapse = ","), "&",
                "format=image/tiff&",
                "interpolation=nearest&resx=0.00833333&resy=0.00833333"
  )
  filename <- file.path('data', 'raster',
                                     paste0(year, '_', month, '_', vessel_code,'.tif'))
  download.file(url = url,
                destfile = filename,
                mode='wb')
  x <- raster(filename)
}

```

We extract the Vessel density data, for example here for the Fishing map of January 2019:
(units are *number of routes / km² / ship type*)
```{r get_WCS_data, message = FALSE, warning = FALSE, echo = FALSE}
# extract vessel density data and store it in a rasterbrick:

Cargo <- brick(x = c('Cargo-2019-01' = getBelgianCSW(2019, 1, "01"),
                     'Cargo-2019-02' = getBelgianCSW(2019, 2, "01"),
                     'Cargo-2019-03' = getBelgianCSW(2019, 3, "01"),
                     'Cargo-2019-04' = getBelgianCSW(2019, 4, "01"),
                     'Cargo-2019-05' = getBelgianCSW(2019, 5, "01"),
                     'Cargo-2019-06' = getBelgianCSW(2019, 6, "01"),
                     'Cargo-2019-07' = getBelgianCSW(2019, 7, "01"),
                     'Cargo-2019-08' = getBelgianCSW(2019, 8, "01"),
                     'Cargo-2019-09' = getBelgianCSW(2019, 9, "01"),
                     'Cargo-2019-10' = getBelgianCSW(2019, 10, "01"),
                     'Cargo-2019-11' = getBelgianCSW(2019, 11, "01"),
                     'Cargo-2019-12' = getBelgianCSW(2019, 12, "01"),
                     'Cargo-2020-01' = getBelgianCSW(2020, 1, "01"),
                     'Cargo-2020-02' = getBelgianCSW(2020, 2, "01"),
                     'Cargo-2020-03' = getBelgianCSW(2020, 3, "01"),
                     'Cargo-2020-04' = getBelgianCSW(2020, 4, "01")
                                   )
)


Fishing <- brick(x = c('Fishing-2019-01' = getBelgianCSW(2019, 1, "02"),
                     'Fishing-2019-02' = getBelgianCSW(2019, 2, "02"),
                     'Fishing-2019-03' = getBelgianCSW(2019, 3, "02"),
                     'Fishing-2019-04' = getBelgianCSW(2019, 4, "02"),
                     'Fishing-2019-05' = getBelgianCSW(2019, 5, "02"),
                     'Fishing-2019-06' = getBelgianCSW(2019, 6, "02"),
                     'Fishing-2019-07' = getBelgianCSW(2019, 7, "02"),
                     'Fishing-2019-08' = getBelgianCSW(2019, 8, "02"),
                     'Fishing-2019-09' = getBelgianCSW(2019, 9, "02"),
                     'Fishing-2019-10' = getBelgianCSW(2019, 10, "02"),
                     'Fishing-2019-11' = getBelgianCSW(2019, 11, "02"),
                     'Fishing-2019-12' = getBelgianCSW(2019, 12, "02"),
                     'Fishing-2020-01' = getBelgianCSW(2020, 1, "02"),
                     'Fishing-2020-02' = getBelgianCSW(2020, 2, "02"),
                     'Fishing-2020-03' = getBelgianCSW(2020, 3, "02"),
                     'Fishing-2020-04' = getBelgianCSW(2020, 4, "02")
                     )
)

mapview(Fishing[[1]])

Passenger <- brick(x = c('Passenger-2019-01' = getBelgianCSW(2019, 1, "03"),
                        'Passenger-2019-02' = getBelgianCSW(2019, 2, "03"),
                        'Passenger-2019-03' = getBelgianCSW(2019, 3, "03"),
                        'Passenger-2019-04' = getBelgianCSW(2019, 4, "03"),
                        'Passenger-2019-05' = getBelgianCSW(2019, 5, "03"),
                        'Passenger-2019-06' = getBelgianCSW(2019, 6, "03"),
                        'Passenger-2019-07' = getBelgianCSW(2019, 7, "03"),
                        'Passenger-2019-08' = getBelgianCSW(2019, 8, "03"),
                        'Passenger-2019-09' = getBelgianCSW(2019, 9, "03"),
                        'Passenger-2019-10' = getBelgianCSW(2019, 10, "03"),
                        'Passenger-2019-11' = getBelgianCSW(2019, 11, "03"),
                        'Passenger-2019-12' = getBelgianCSW(2019, 12, "03"),
                        'Passenger-2020-01' = getBelgianCSW(2020, 1, "03"),
                        'Passenger-2020-02' = getBelgianCSW(2020, 2, "03"),
                        'Passenger-2020-03' = getBelgianCSW(2020, 3, "03"),
                        'Passenger-2020-04' = getBelgianCSW(2020, 4, "03")
)
)

Tanker <- brick(x = c('Tanker-2019-01' = getBelgianCSW(2019, 1, "04"),
                      'Tanker-2019-02' = getBelgianCSW(2019, 2, "04"),
                      'Tanker-2019-03' = getBelgianCSW(2019, 3, "04"),
                      'Tanker-2019-04' = getBelgianCSW(2019, 4, "04"),
                      'Tanker-2019-05' = getBelgianCSW(2019, 5, "04"),
                      'Tanker-2019-06' = getBelgianCSW(2019, 6, "04"),
                      'Tanker-2019-07' = getBelgianCSW(2019, 7, "04"),
                      'Tanker-2019-08' = getBelgianCSW(2019, 8, "04"),
                      'Tanker-2019-09' = getBelgianCSW(2019, 9, "04"),
                      'Tanker-2019-10' = getBelgianCSW(2019, 10, "04"),
                      'Tanker-2019-11' = getBelgianCSW(2019, 11, "04"),
                      'Tanker-2019-12' = getBelgianCSW(2019, 12, "04"),
                      'Tanker-2020-01' = getBelgianCSW(2020, 1, "04"),
                      'Tanker-2020-02' = getBelgianCSW(2020, 2, "04"),
                      'Tanker-2020-03' = getBelgianCSW(2020, 3, "04"),
                      'Tanker-2020-04' = getBelgianCSW(2020, 4, "04")
)
)

Other <- brick(x = c('Other-2019-01' = getBelgianCSW(2019, 1, "05"),
                      'Other-2019-02' = getBelgianCSW(2019, 2, "05"),
                      'Other-2019-03' = getBelgianCSW(2019, 3, "05"),
                      'Other-2019-04' = getBelgianCSW(2019, 4, "05"),
                      'Other-2019-05' = getBelgianCSW(2019, 5, "05"),
                      'Other-2019-06' = getBelgianCSW(2019, 6, "05"),
                      'Other-2019-07' = getBelgianCSW(2019, 7, "05"),
                      'Other-2019-08' = getBelgianCSW(2019, 8, "05"),
                      'Other-2019-09' = getBelgianCSW(2019, 9, "05"),
                      'Other-2019-10' = getBelgianCSW(2019, 10, "05"),
                      'Other-2019-11' = getBelgianCSW(2019, 11, "05"),
                      'Other-2019-12' = getBelgianCSW(2019, 12, "05"),
                      'Other-2020-01' = getBelgianCSW(2020, 1, "05"),
                      'Other-2020-02' = getBelgianCSW(2020, 2, "05"),
                      'Other-2020-03' = getBelgianCSW(2020, 3, "05"),
                      'Other-2020-04' = getBelgianCSW(2020, 4, "05")
)
)

All <- brick(x = c('All-2019-01' = getBelgianCSW(2019, 1, "All"),
                   'All-2019-02' = getBelgianCSW(2019, 2, "All"),
                   'All-2019-03' = getBelgianCSW(2019, 3, "All"),
                   'All-2019-04' = getBelgianCSW(2019, 4, "All"),
                   'All-2019-05' = getBelgianCSW(2019, 5, "All"),
                   'All-2019-06' = getBelgianCSW(2019, 6, "All"),
                   'All-2019-07' = getBelgianCSW(2019, 7, "All"),
                   'All-2019-08' = getBelgianCSW(2019, 8, "All"),
                   'All-2019-09' = getBelgianCSW(2019, 9, "All"),
                   'All-2019-10' = getBelgianCSW(2019, 10, "All"),
                   'All-2019-11' = getBelgianCSW(2019, 11, "All"),
                   'All-2019-12' = getBelgianCSW(2019, 12, "All"),
                   'All-2020-01' = getBelgianCSW(2020, 1, "All"),
                   'All-2020-02' = getBelgianCSW(2020, 2, "All"),
                   'All-2020-03' = getBelgianCSW(2020, 3, "All"),
                   'All-2020-04' = getBelgianCSW(2020, 4, "All")
)
)

```

```{r writeraster, message = FALSE, warning = FALSE, echo = FALSE}
if (!dir.exists("output")) dir.create("output")

names <- c('Cargo', 'Fishing', 'Passenger', 'Tanker', 'Other', 'All')
bricks <- c(Cargo, Fishing, Passenger, Tanker, Other, All)

l <- lapply(bricks,
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)

```



### Belgian EEZ:

We extract all data from all the cells, and calculate the average for the whole EEZ.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# No we create a data frame to store the statistics that we will extract from the rasters.
# We extract for all values of the Belgian EEZ, the average value of all raster cells.

# create data frame, with 1st column 'time'
  df.sr <- data.frame(Date = seq(as.Date("2019/1/1"), by = "month", length.out = 16))

  # extract all values from the rasters, by the BCP polygon, and average the values.
  df.sr$Cargo     <- as.vector(extract(Cargo, BCP, fun = mean))
  df.sr$Fishing   <- as.vector(extract(Fishing, BCP, fun = mean))
  df.sr$Passenger <- as.vector(extract(Passenger, BCP, fun = mean))
  df.sr$Tanker    <- as.vector(extract(Tanker, BCP, fun = mean))
  df.sr$Other     <- as.vector(extract(Other, BCP, fun = mean))
  df.sr$All       <- as.vector(extract(All, BCP, fun = mean))

# transform to long format:
  df.sr.long <- melt.data.table(as.data.table(df.sr),
                                id.vars = "Date")
```
The table below shows the average values for the whole Belgian EEZ, for the different fishing types. We only have monthly data, so each data represents one month, from January 2019 to April 2020.
(units are: *number of routes / km² / ship type*)

```{r BCP_table_wide, echo = FALSE}
datatable(df.sr,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)
```

```{r BCP_table_long, echo = FALSE, eval = FALSE}
datatable(df.sr.long,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(3, 2)
```

A plot of the data:

```{r plot, message = FALSE, warning = FALSE, echo = FALSE}

ggplot(df.sr.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Vessel route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()
```


### Scheldt estuary:

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# create data frame, with 1st column 'time'
  df.Scheldt <- data.frame(Date = seq(as.Date("2019/1/1"), by = "month", length.out = 16))

  # extract all values from the rasters, by the WScheldt polygon, and average the values.
  df.Scheldt$Cargo     <- as.vector(extract(Cargo, WScheldt, fun = mean))
  df.Scheldt$Fishing   <- as.vector(extract(Fishing, WScheldt, fun = mean))
  df.Scheldt$Passenger <- as.vector(extract(Passenger, WScheldt, fun = mean))
  df.Scheldt$Tanker    <- as.vector(extract(Tanker, WScheldt, fun = mean))
  df.Scheldt$Other     <- as.vector(extract(Other, WScheldt, fun = mean))
  df.Scheldt$All       <- as.vector(extract(All, WScheldt, fun = mean))

# transform to long format:
  df.Scheldt.long <- melt.data.table(as.data.table(df.Scheldt),
                                id.vars = "Date")

```
The table below shows the average values for the Western Scheldt, for the different fishing types. We only have monthly data, so each data represents one month, from January 2019 to April 2020.
(units are *number of routes / km² / ship type*)
```{r Scheldt table, echo = FALSE}
datatable(df.Scheldt,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)
```


A plot of the result:

```{r Schelde_plot, message = FALSE, warning = FALSE, echo = FALSE}

ggplot(df.Scheldt.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Vessel route densities in the Western Scheldt",
       subtitle = "Monthly average route density (routes/km²/ship type) in the Western Scheldt",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()
```


## 2020 vs 2019

We create a average value for the periods

 - February - April 2019
 - February - April 2020
 
And plot both periods. For each vessel type, both periods are visualised on a single map.
(units are *number of routes / km² / ship type*)

```{r message = FALSE, warning = FALSE, echo = FALSE}
# Average map of February-April 2019
r2019_0204 <- brick(c(
  Cargo_2019_0204 = mean(Cargo[[2:4]]),
  Fishing_2019_0204 = mean(Fishing[[2:4]]),
  Passenger_2019_0204 = mean(Passenger[[2:4]]),
  Tanker_2019_0204 = mean(Tanker[[2:4]]),
  Other_2019_0204 = mean(Other[[2:4]]),
  All_2019_0204 = mean(All[[2:4]])
)
)
  
# Average map of February-April 2020
r2020_0204 <- brick(c(
  Cargo_2020_0204 = mean(Cargo[[14:16]]),
  Fishing_2020_0204 = mean(Fishing[[14:16]]),
  Passenger_2020_0204 = mean(Passenger[[14:16]]),
  Tanker_2020_0204 = mean(Tanker[[14:16]]),
  Other_2020_0204 = mean(Other[[14:16]]),
  All_2020_0204 = mean(All[[14:16]])
)
)


l <- lapply(list(r2019_0204,r2020_0204),
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)

```

```{r mapviewpolygons, message = FALSE, warning = FALSE, echo = FALSE}
mapviewpolygons <- 
  mapview(BCP, color = 'blue',
        alpha.regions = 0,
        legend = FALSE,
        label = NULL,
        popup = NULL,
        homebutton = FALSE) +
  mapview(WScheldt,
          color = 'blue',
          alpha.regions = 0,
          legend = FALSE,
          label = NULL,
          popup = NULL,
        homebutton = FALSE) +
  mapview(B3NM,
          legend = FALSE, 
          label = NULL,
          popup = NULL,
        homebutton = FALSE) +
  mapview(B12NM,
          legend = FALSE, 
          label = NULL,
          popup = NULL,
        homebutton = FALSE)

```


### Cargo 

Cargo ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Cargo_2019_0204) + 
  mapview(r2020_0204$Cargo_2020_0204) 

```

### Fishing 

Fishing ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Fishing_2019_0204) + 
  mapview(r2020_0204$Fishing_2020_0204)
```

### Passenger 

Passenger ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020
mapviewpolygons +
  mapview(r2019_0204$Passenger_2019_0204) + 
  mapview(r2020_0204$Passenger_2020_0204)
```

### Tanker 

Tanker ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020
mapviewpolygons +
  mapview(r2019_0204$Tanker_2019_0204) + 
  mapview(r2020_0204$Tanker_2020_0204)
```

### Other 

Other ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$Other_2019_0204) + 
  mapview(r2020_0204$Other_2020_0204)
```

### All 

All ships Feb-April 2019 vs Feb-April 2020 (units are *number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the February - April Fishing maps for 2019 and 2020

mapviewpolygons +
  mapview(r2019_0204$All_2019_0204) + 
  mapview(r2020_0204$All_2020_0204)
```


## Anomaly maps

### February - April 2019 compared to February - April 2020
Comparing the period February-April 2020 vs 2019:
The different layers are the different boat types, click on the 'layers' button to visualise different layers.
(units are the difference in *number of routes / km² / ship type*)


```{r echo = FALSE}
anomaly_February_April <- r2020_0204 - r2019_0204
names(anomaly_February_April) <- c("Anomaly_Cargo_Feb_Apr_2019_2020",
                                   "Anomaly_Fishing_Feb_Apr_2019_2020",
                                   "Anomaly_Passenger_Feb_Apr_2019_2020",
                                   "Anomaly_Tanker_Feb_Apr_2019_2020",
                                   "Anomaly_Other_Feb_Apr_2019_2020",
                                   "Anomaly_All_Feb_Apr_2019_2020") 
writeRaster(anomaly_February_April,
            gsub('\\.','_',file.path('output', names(anomaly_February_April))),
            format = "GTiff",
            bylayer = TRUE,
            overwrite = TRUE)


pal <- grDevices::colorRampPalette(grDevices::hcl.colors(255, palette = "BrBG"))

mapviewpolygons +
  mapView(anomaly_February_April,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )

```


### Month-by-month comparisons 2019 vs 2020
Comparing the month-by-month differences between 2020 vs 2019:
You can select the different layers by clicking on the layer symbol.


```{r message = FALSE, warning = FALSE, echo = FALSE}
# Anomaly map January-April 2020 - January-April 2019
Anomaly_Cargo <- brick(c(
  Anomaly_Cargo_January = mean(Cargo[[13]] - Cargo[[1]]),
  Anomaly_Cargo_February = mean(Cargo[[14]] - Cargo[[2]]),
  Anomaly_Cargo_March = mean(Cargo[[15]] - Cargo[[3]]),
  Anomaly_Cargo_April = mean(Cargo[[16]] - Cargo[[4]])
)
)

Anomaly_Fishing <- brick(c(
  Anomaly_Fishing_January = mean(Fishing[[13]] - Fishing[[1]]),
  Anomaly_Fishing_February = mean(Fishing[[14]] - Fishing[[2]]),
  Anomaly_Fishing_March = mean(Fishing[[15]] - Fishing[[3]]),
  Anomaly_Fishing_April = mean(Fishing[[16]] - Fishing[[4]])
)
)

Anomaly_Passenger <- brick(c(
  Anomaly_Passenger_January = mean(Passenger[[13]] - Passenger[[1]]),
  Anomaly_Passenger_February = mean(Passenger[[14]] - Passenger[[2]]),
  Anomaly_Passenger_March = mean(Passenger[[15]] - Passenger[[3]]),
  Anomaly_Passenger_April = mean(Passenger[[16]] - Passenger[[4]])
)
)

Anomaly_Tanker <- brick(c(
  Anomaly_Tanker_January = mean(Tanker[[13]] - Tanker[[1]]),
  Anomaly_Tanker_February = mean(Tanker[[14]] - Tanker[[2]]),
  Anomaly_Tanker_March = mean(Tanker[[15]] - Tanker[[3]]),
  Anomaly_Tanker_April = mean(Tanker[[16]] - Tanker[[4]])
)
)

Anomaly_Other <- brick(c(
  Anomaly_Other_January = mean(Other[[13]] - Other[[1]]),
  Anomaly_Other_February = mean(Other[[14]] - Other[[2]]),
  Anomaly_Other_March = mean(Other[[15]] - Other[[3]]),
  Anomaly_Other_April = mean(Other[[16]] - Other[[4]])
)
)

Anomaly_All <- brick(c(
  Anomaly_All_January = mean(All[[13]] - All[[1]]),
  Anomaly_All_February = mean(All[[14]] - All[[2]]),
  Anomaly_All_March = mean(All[[15]] - All[[3]]),
  Anomaly_All_April = mean(All[[16]] - All[[4]])
)
)


l <- lapply(list(Anomaly_Cargo, Anomaly_Fishing, Anomaly_Passenger,
            Anomaly_Tanker, Anomaly_Other, Anomaly_All
            ),
       function(x){
          writeRaster(x,
                      gsub('\\.','_',file.path('output', names(x))),
                      format = "GTiff",
                      bylayer = TRUE,
                      overwrite = TRUE)
       }
)


```

### Cargo 

Anomaly of cargo ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Cargo,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Fishing 

Anomaly of fishing ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Fishing,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Passenger 

Anomaly of passenger ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Passenger,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Tanker 

Anomaly of tanker ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Tanker,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### Other 

Anomaly of other ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_Other,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

### All 

Anomaly of all ships per month (units are *difference in number of routes / km² / ship type*)
```{r, message = FALSE, warning = FALSE, echo = FALSE}
# Plot the monthtly anomaly maps for 2019 vs 2020
mapviewpolygons +
  mapView(Anomaly_All,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )
```

## Additional analyses

### Passengers excluding wind farm

There is quite some traffic to the windfarms that are classified as 'Passenger' traffic.
As this might be confusing, we will exclude this from the analysis. The excluded area is visualised in blue on the map below.

```{r, message = FALSE, warning = FALSE, echo = FALSE}
passenger_windfarm <- st_read('data/passenger_windfarm.gpkg', quiet = TRUE)

mapviewpolygons +
  mapview(passenger_windfarm,
          label = NULL,
          popup = NULL,
          homebutton = FALSE) +
  mapview(Anomaly_Passenger,
        col.regions = pal,
        at = c(-1000, -200, -100, -50, -25,
               25, 50, 100, 200, 1000)
        )

```


```{r, message = FALSE, warning = FALSE, echo = FALSE}

BCP_no_windfarm <- st_difference(BCP, passenger_windfarm)

df.sr$Pass_no_windfarm <- as.vector(extract(Passenger, BCP_no_windfarm, fun = mean))

df.sr.pass <- df.sr[c('Date','Passenger','Pass_no_windfarm')]

# transform to long format:
  df.sr.pass.long <- melt.data.table(as.data.table(df.sr.pass),
                                id.vars = "Date")


```
Below is a table showing the average monthly passenger density ('Passenger' column) and the same analysis but excluding the traffic to the wind farm ('Pass_no_windfarm').

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
datatable(df.sr.pass,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)

ggplot(df.sr.pass.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Passenger route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "Ship type",
       shape = "Ship type") +
  theme_minimal()

```

### Fishing zones.

In this part we'll look how much is being fished in the different zones within the Belgian EEZ:

* from the coastline to 3 nautical miles (NM)
* from 3 to 12 nautical miles (NM)
* further than 12 nautical miles from the coastline

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
# extended the lines of 3 and 12 NM a bit
BCP_zones <- st_read('data/eez_be_poly_split.gpkg', quiet = TRUE)
BCP_zones$name <- c(">12 NM", "0-3 NM", "3-12 NM")
mapview(BCP_zones)

#feb-apr20 vs feb-apr19 Fishing
numberofcells <- extract(Fishing, BCP_zones,
                         fun = function(x, ...) length(x),
                         df = TRUE)

df.Fishing.zones <- extract(Fishing, BCP_zones, fun = mean, df = TRUE)
df.Fishing.zones <- transpose(df.Fishing.zones[-1])
colnames(df.Fishing.zones) <- BCP_zones$name
df.Fishing.zones$Date <- seq(as.Date("2019/1/1"), by = "month", length.out = 16)

df.Fishing.zones <- df.Fishing.zones[c("Date",  "0-3 NM", "3-12 NM", ">12 NM")]

# transform to long format:
  df.Fishing.zones.long <- melt.data.table(as.data.table(df.Fishing.zones),
                                id.vars = "Date")


```

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
datatable(df.Fishing.zones,
          options = list(pageLength = 16),
          rownames = FALSE) %>% formatRound(-1, 2)

ggplot(df.Fishing.zones.long, aes(x = Date, y = value, group = variable)) +
  geom_point(aes(color = variable, shape = variable)) +
  geom_line(aes(color = variable)) +
  labs(title = "Fishing route densities in the Belgian part of the North Sea",
       subtitle = "Monthly average route density (routes/km²/ship type) for the Belgian EEZ",
       x = "Month",
       y = "Average routes / km² / ship type",
       color = "zone",
       shape = "zone") +
  theme_minimal()


```


## Data availability

- *Vessel densities*
The data from this exercise is freely available at the [EMODnet Human activities](https://www.emodnet-humanactivities.eu/) portal. [EMODnet](https://www.emodnet.eu/) stands for the European Marine Observation and Data Network and is a network of organisations that are collecting and standardizing European marine data, and making those data products freely available, supported by EU's integrated marine policy.
The vessel densities data used in this exercise are provided by the European Maritime Safety Agency (EMSA) to EMODnet human activities and are available [here](https://www.emodnet-humanactivities.eu/search-results.php?dataname=Vessel+Density+).

- *Maritime boundaries*
The maritime boundaries used in this exercise are from [MarineRegions.org](http://www.marineRegions.org). MarineRegions.org maintains a standard, relational list of geographic names coupled with information and maps of the geographic location of these features. This improves access and clarity of the different geographic, marine names and allows an improved linking of these locations to databases. Marine Regions is developed by Flanders Marine Institute (VLIZ) as part of the Flemish contribution to [LifeWatch](http://www.lifewatch.be/), funded by [Research Foundation - Flanders](https://www.fwo.be/en/).
The more information about the polygons used in this exercise:
  - the Western Scheldt: http://marineregions.org/gazetteer.php?p=details&id=4752
  - the Belgian Exclusive Economic Zone: http://marineregions.org/gazetteer.php?p=details&id=3293

## Code availability

All the code needed to run this analysis is available [here](https://github.com/vlizBE/Vessel_route_densities/).

This code makes use of following R packages:

- raster: for raster data
- sf: for spatial data
- mapview: for interactive maps
- ggplot2: for plots
- data.table: for manipulation dataframes/tables
- DT: for visualisation of the data tables
- mregions: for standardize marine regions from http://www.marineregions.org

## Reproducibility

You can re-run the full analysis in an RStudio environment by clicking on the button below.

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/vlizBE/Vessel_route_densities/master?urlpath=rstudio)

This will open an online RStudio environment. To run the analysis, open the 'index.Rmd' file.
In this file, you can run separate code chunks or click on the 'knit' button to recreate the html file
(this will take a while, and open a pop-up window).
You can also edit the code and run your own analysis.

If you have any issues with running the analysis, please let us know by [opening an issue](https://github.com/vlizBE/Vessel_route_densities/issues).


